---
title: "Shooting Data"
author: "Josh Garvey"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Step 1: Obtaining Data
First thing I will do is read in the shooting data from the web.
```{r get_data, include=TRUE}
url_in <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
shooting_data <- read_csv(url_in)
```

## Step 2: Tidy Up the data
Now that I have data in the shooting_data variable I can work on tidying it up. I would like to look at 
the amount of deaths/shooting/precinct. I left in victim and perpetrator sex and race in case I want to 
start pulling those factors in as well
```{r tidy_data, include=TRUE}
sub_shoot = subset(shooting_data, select = -c(INCIDENT_KEY, 
                                              OCCUR_DATE,
                                              OCCUR_TIME, 
                                              BORO, 
                                              JURISDICTION_CODE, 
                                              LOCATION_DESC, 
                                              X_COORD_CD, 
                                              Y_COORD_CD,
                                              VIC_AGE_GROUP,
                                              Latitude, 
                                              Longitude, 
                                              Lon_Lat
                                              ))
# columns removed
sub_shoot
# summary
summary(sub_shoot)
```
There are some null points in the data set. I plan on throwing those data points out
if they are part of an analysis.


## Step 3: Visualize the Data and Analyze
With the data parsed down I can start to explore different factors leading to 
shooting deaths to try and find correlations. To start with I will select the
columns for whether or not there was a death, and what precinct it occured in.
```{r vis_data, include=TRUE}
deaths = subset(sub_shoot, select = c(PRECINCT, STATISTICAL_MURDER_FLAG))
summary(deaths)

# Total shootings and percentage that are lethal (a baseline for the area)
nLethal = sum(!deaths[2])
Lethal = sum(deaths[2])
total_shootings = Lethal + nLethal
percent_lethal_total = (Lethal/total_shootings)*100

# Group the data to be graphed
deaths_by_precinct <- deaths %>%
  group_by(PRECINCT) %>%
  summarize(num_deaths = sum(STATISTICAL_MURDER_FLAG), 
            shootings = sum(STATISTICAL_MURDER_FLAG)+sum(!STATISTICAL_MURDER_FLAG)) %>%
  mutate(perc_lethal = (num_deaths / shootings) *100) %>%
  ungroup()

# Graph the lethal percentages by precinct
deaths_by_precinct %>%
  ggplot(aes(x = PRECINCT, y = perc_lethal)) +
  geom_line(aes(color = "precinct lethality")) +
  geom_point(aes(color = "precinct lethality")) +
  geom_line(aes(y = percent_lethal_total, color = "city average")) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "Lethal Shootings By Precinct", y = "Percentage")

# Graph number of shootings versus total shooting to see if there is a trigger happy precinct
shootings_by_precinct <- deaths %>%
  group_by(PRECINCT) %>%
  summarize(num_shootings = sum(STATISTICAL_MURDER_FLAG)+sum(!STATISTICAL_MURDER_FLAG)) %>%
  ungroup()

shootings_by_precinct %>%
  ggplot(aes(x = PRECINCT, y = num_shootings)) +
  geom_line(aes(color = "shootings")) +
  geom_point(aes(color = "shootings")) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "Shootings By Precinct", y = NULL)

```

## Step 4: Sources of Bias
I think the biggest bias source is the fact that I chose to include factors like 
race and sex into my analysis but left out age range, location, the date, etc. There
are infinitely many factors that go into an officers decision to shoot and by 
handpicking these ones to analyze I may be missing some huge correlation. An example
could be that the rate of deaths by shooting goes way up in one location due to it
being further from medical attention than another. 

I could try to mitigate by picking a few columns for data to hopefully give me an idea
when something is an outlier. This is going to be hard to pick arbitrarily. Alternatively,
I can accept that there is bias in the study and append a synopsis of how I got
my results to the report so others can decide if its worthy of their use.

```{r session, include=FALSE}
sessionInfo()
```